\documentclass{article}
\usepackage{fontspec}                     %設定字體用
\setmainfont{新細明體}                     %使用 Windows 系統中文字體(標楷體, 新細明體)
\XeTeXlinebreaklocale "zh"                %讓中文自動換行
\XeTeXlinebreakskip = 0pt plus 1pt        %這兩行是texWorks範例中所缺少的。

\title{採用 TexStudio}
\author{作者姓名}
\date{} %不要日期

\begin{document}
\maketitle

\section {開始使用 XeLatex}
This is a very  simple document.

這是我的中文內容，也可以直接打公式以及符號，\fontspec{標楷體}這裡開始改用標楷體。這是我的中文內容，也可以直接打公式以及符號，這裡也是中文內容。這是我的中文內容，也可以直接打公式以及符號，這裡也是中文內容。這是我的中文內容，也可以直接打公式以及符號，這裡也是中文內容。這是我的中文內容，也可以直接打公式以及符號，這裡也是中文內容。這是我的中文內容，也可以直接打公式以及符號，這裡也是中文內容。這是我的中文內容，也可以直接打公式以及符號，\fontspec{新細明體}這裡開始改用細明體。這是我的中文內容，也可以直接打公式以及符號，這裡也是中文內容。這是我的中文內容，也可以直接打公式以及符號，這裡也是中文內容。這是我的中文內容，也可以直接打公式以及符號，這裡也是中文內容。

English Test.  This is a simple template for a XeLaTeX document using the
article class, with the fontspec package to easily select fonts.
Don't forget to include examples of topicalization.
They look like this:

\section {一般文字}
這裡當然可以處理許多的字元,也沒有問題,但是在許多情況下也並非沒有
Type your text in free-format; lines can be as long
or as short
as you wish.
      You can indent      or space out
        your input
          text in
            any way you like to highlight the structure
      of your manuscript and make it easier to edit.
LaTeX fills lines and adjusts spacing between words to produce an
aesthetically pleasing result.

Completely blank lines in the input file break your text into
paragraphs.
To change the font for a single character, word, or set of words,
enclose the word and the font changing command within braces,
{\em like this}.
A font changing command not enclosed in braces, like the change to \bf
bold here, keeps that change in effect until the end of the document or
until countermanded by another font switch, like this change back to
\rm roman. 


\section {顯示文字}
Use the ``quote'' and ``quotation'' environments for typesetting quoted
material or any other text that should be slightly indented and set off
from the normal text.
\begin{quotation}
The quote and quotation environments are similar, but use different
settings for paragraph indentation and spacing.

\em When in doubt, consult the manual.
\end{quotation}

So far, I have demonstrated titles, paragraphs, font changes, and
section headings.
Now, I am going to show lists and tables. 

\begin{enumerate}
\item
這是第一項
\item 
這是第二項
\item
這是第三項

$a^2+b^2=c^2$

\end{enumerate}

Items in a list can contain multiple paragraphs.
These paragraphs are appropriately spaced and indented according to their
position in the list.
  \begin{itemize}
  \item The ``itemize'' environment sets off list items with ``bullets'',
like this. Finally, the ``description'' environment lets you put your own
    \begin{description}
    \item[A] label on each item, like this ``A''.
    \item[If the label is long,] the first line of the item text will
be spaced over to the right as needed.
    \end{description} 
  \end{itemize}

Here is a very simple table showing data lined up in columns.
Notice that I include the table in a ``center'' environment to display
it properly.
The title is created simply as another paragraph in the center environment,
rather than as part of the table itself.

\begin{center}
Numbers of Computers on Earth Sciences Network, By Type.

\begin{tabular}{lr}
Macintosh&175\\
DOS/Windows PC&60\\
Unix Workstation or server&110\\
\end{tabular}
\end{center} 


Here is a more complicated table that has been boxed up, with a multi-column
header and paragraph entries set in one of the columns.
\begin{center}
\begin{tabular}{|l|c|p{3.5in}|}
\hline
\multicolumn{3}{|c|}{Places to Go Backpacking}\\ \hline
Name&Driving Time&Notes\\
&(hours)&\\ \hline
Big Basin&1.5&Very nice overnight to Berry Creek Falls from
either Headquarters or ocean side.\\ \hline
Sunol&1&Technicolor green in the spring. Watch out for the cows.\\ \hline
Henry Coe&1.5&Large wilderness nearby suitable for multi-day treks.\\ \hline
\end{tabular}
\end{center} 

\section {數學方程式}
Simple equations, like $x^y$ or $x_n = \sqrt{a + b}$ can be typeset right
in the text line by enclosing them in a pair of single dollar sign symbols.
Don't forget that if you want a real dollar sign in your text, like \$2000,
you have to use the \verb+\$+ command.

A more complicated equation should be typeset in {\em displayed math\/} mode,
like this:
\[
z \left( 1 \ +\ \sqrt{\omega_{i+1} + \zeta -\frac{x+1}{\Theta +1} y + 1}
\ \right)
\ \ \ =\ \ \ 1
\]
The ``equation'' environment displays your equations, and automatically
numbers them consecutively within your document, like this:
\begin{equation}
\left[
{\bf X} + {\rm a} \ \geq\
\underline{\hat a} \sum_i^N \lim_{x \rightarrow k} \delta C
\right]
\end{equation}

\begin{equation}
\left[
{\bf X} + {\rm a} \ \geq\
\underline{\hat a} \sum_i^N \lim_{x \rightarrow k} \delta C
\right]
\end{equation}

\end{document}