% This file is part of the "lettre" package.
% This package is distributed under the terms of the LaTeX Project 
% Public License (LPPL) described in the file lppl.txt.
%
% Denis Mégevand - Observatoire de Genève.
%
% Ce fichier fait partie de la distribution du paquetage "lettre".
% Ce paquetage est distribué sous les termes de la licence publique 
% du projet LaTeX (LPPL) décrite dans le fichier lppl.txt.

\documentclass[12pt,origdate]{lettre}
\usepackage[francais,english]{babel}
\usepackage[OT1]{fontenc}
\usepackage{mltex}
\begin{document}
%
% Fichier de defaut de l'Observatoire
% ===================================
\institut{obs}
%
% Entete par defaut, double signature,
% ====================================
% langue anglaise, format headings, marge precisee.
% =================================================
% Champs: objet.
% ==============
% Lignes auxiliaires de reference et d'E-Mail. 
% ============================================
% Tests: itemize, description, enumerate, center, 
% ===============================================
%        tabular, array, eqnarray, footnote.
%        ===================================
%
\begin{letter}{	SoftIdeas\\
                P.O.Box 8K23 \\
                Prove Islands, TO 14501 \\
                USA}

\anglais
\pagestyle{headings}
\marge{30mm}

\name{Dr~S.~E.~Dnavegem}
\signature{Dr~S.~E.~Dnavegem\\Collaborateur Scientifique}
\secondsignature{Dr~P.~Barre-Thordue\\Collaborateur Scientifique}

\nref{PL/cf69}
\faxobs
\username{dnavegem}
\ccittobs
\internetobs

\conc{Our activities} 

\opening{Dear Sir,}

Enclosed\footnote{In fact, the check will be sent by separate mail}, you 
will find a description of our activities and a check of \$149.95 corresponding 
to your invoice \#439 dated 05/28/93 for:
\begin{itemize}
\item 1 PiC\TeX\ manual, describing completely all the commands implemented
by the macros written by M.J.~Wichura. If the manual doesn't respond to
our demand, we will send it back, ans ask for complete refund.
\item 1 PiC\TeX\ software on 5.25 inch floppy disk. The disk is to be
formatted in IBM PC compatible format.
\item 5 PiC\TeX\ Quick Reference Cards.
\end{itemize}
We'll be happy if you could send us a detailled list of your products.

We are a software developing team, and our work involves:
\begin{description}
\item[Editors] Full-screen editors for PC and compatibles.
\item[Compilers] Full range of compilers for PC and compatibles. It includes
ADA, BASIC, C, FORTH, FORTRAN, ICON, LISP, MODULA-2, PASCAL and PROLOG 
compilers.
\item[Operating Systems] Multitasking, real-time OS.
\item[Utilities] A lot of useful add-on software.
\end{description}

We can produce cross-reference tables of product compatibilities for various
software and vendors.
\begin{center}
\begin{tabular}{||l|||r@{.}l|l|l||}
\hline
gnuBASE IV&419&00&nul&ok\\
\hline
GNATUS 123&489&95&complete&N.A\\
\hline
Turbo Foo&99&95&partial&good\\
\hline
\end{tabular}
\end{center}

We can also furnish scientific text processing systems for your computer,
or for your site (Education licenses!)
Here are some examples of our outputs:

\begin{eqnarray}
x&=&17y\\
y&>&a+b+c+d+e+f+g+h+i+j+ \nonumber \\
&&k+l+m+n+o+p
\end{eqnarray}

\[
x=\left\{\begin{array}{ll}
                y&\mbox{if $y>0$} \\
                z+y&\mbox{otherwise}
         \end{array}
  \right.
\]

{\em et voila \ldots that's it \ldots}

Remember we order today:
\begin{enumerate}
\item 1 PiC\TeX\ manual, describing completely all the commands implemented
by the macros written by M.J.~Wichura. 
\item 1 PiC\TeX\ software on 5.25 inch IBM PC compatible formatted floppy disk. 
\item 5 PiC\TeX\ Quick Reference Cards.
\end{enumerate}

\closing{Sincerely}

\end{letter}
%
\end{document}
