%%
%% This is file `lastpage-example.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% lastpage.dtx  (with options: `example')
%% 
%% This is a generated file.
%% 
%% Project: lastpage
%% Version: 2011/08/08 v1.2i
%% 
%% Copyright (C) 2010, 2011 by
%%     H.-Martin M"unch <Martin dot Muench at Uni-Bonn dot de>
%% Portions of code copyrighted by other people as marked.
%% 
%% The usual disclaimer applys:
%% If it doesn't work right that's your problem.
%% (Nevertheless, send an e-mail to the maintainer
%%  when you find an error in this package.)
%% 
%% This work may be distributed and/or modified under the
%% conditions of the LaTeX Project Public License, either
%% version 1.3c of this license or (at your option) any later
%% version. This version of this license is in
%%    http://www.latex-project.org/lppl/lppl-1-3c.txt
%% and the latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.3c or later is part of all distributions of
%% LaTeX version 2005/12/01 or later.
%% 
%% This work has the LPPL maintenance status "maintained".
%% 
%% The Current Maintainer of this work is H.-Martin Muench.
%% 
%% This package was invented by Jeffrey P. Goldberg.
%% I thought that a replacement was needed and therefore created the
%% pageslts package,
%% http://www.ctan.org/tex-archive/macros/latex/contrib/pageslts/pageslts.dtx
%% or
%% http://www.ctan.org/tex-archive/install/macros/latex/contrib/pageslts.tds.zip
%% . Nevertheless, for compatibility with existing documents/packages as well as
%% for the low amount of resources needed by the lastpage package (no new counter!),
%% I updated this package.
%% Thanks go to Jeffrey P. Goldberg for allowing me to do this.
%% 
%% This work consists of the main source file lastpage.dtx
%% and the derived files
%%    lastpage.sty, lastpage.pdf, lastpage.ins, lastpage.drv,
%%    lastpage-example.tex.
%% 
\documentclass[british]{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\AtEndDocument{\message{^^JLaTeX Info: Executing hook `AtEndDocument'.}}
\usepackage[draft]{showkeys}[2007/08/07]% v3.15
%%      Use final instead of draft to hide the keys. %%
\usepackage{hyperref}[2011/04/17]% v6.82g
\hypersetup{%
 extension=pdf,%
 plainpages=false,%
 pdfpagelabels=true,%
 hyperindex=false,%
 pdflang={en},%
 pdftitle={lastpage package example},%
 pdfauthor={Hans-Martin Muench},%
 pdfsubject={Example for the lastpage package},%
 pdfkeywords={LaTeX, lastpage, Hans-Martin Muench},%
 pdfview=Fit,%
 pdfstartview=Fit,%
 pdfpagelayout=SinglePage%
}
\usepackage{lastpage}[2011/08/08]% v1.2i
\makeatletter
\renewcommand{\@evenfoot}{%
 \normalsize\slshape \today\hfil \upshape page \thepage%
 \ of \pageref{LastPage}}
\renewcommand{\@oddfoot}{\@evenfoot}
\makeatother
\gdef\unit#1{\mathord{\thinspace\mathrm{#1}}}%
\listfiles
\begin{document}
\pagenumbering{Roman}

\section*{Example for lastpage}
\markboth{Example for lastpage}{Example for lastpage}

This example demonstrates the use of package\newline
\textsf{lastpage}, v1.2i as of 2011/08/08 (HMM; JPG).\newline
The package takes no options.\newline
For more details please see the documentation!\newline

\noindent \label{keys} To hide the \pageref{keys}{\qquad } use option
\texttt{final} instead of \texttt{draft} with the \textsf{showkeys}
package (or remove the package call from the preamble of
this document).\newline

\textbf{Hyperlinks or not:} If the \textsf{hyperref} package is loaded,
the references are also hyperlinked:\newline
\smallskip
Last page's name (LastPage): \pageref{LastPage}\newline
\noindent If the \textsf{hyperref} package is loaded, but the hyperlinks
of the references shall be suppressed, \verb|\pageref*{...}|
can be used:\newline
\smallskip
Last page's name (LastPage): \pageref*{LastPage}\newline

\textbf{Trademarks} appear throughout this example without any
trademark symbol; they are the property of their respective
trademark owner. There is no intention of infringement; the
usage is to the benefit of the trademark owner.\newline

\textbf{Tip}: Use \textit{logical page numbers} for
the display of the pdf!\newline
(In Adobe Reader X (10.1.0): \underline{E}dit $>$
Prefere\underline{n}ces (Ctrl+k) $>$ Page Display $>$
Page Content and Information $>$ Use logical page
\nolinebreak{\underline{n}umbers.)}\newline

If you are more ambitiouse in respect to your aims with this package,
you might want to have a look at the \textsf{pageslts} package\newline
\makeatletter
 \@ifundefined{Hy@Warning}{% hyperref not loaded
     \verb|http://www.ctan.org/tex-archive/macros/latex/contrib/pageslts/|).
   }{% hyperref loaded
     \href{http://www.ctan.org/tex-archive/macros/latex/contrib/pageslts/}{%
     CTAN:/macros/latex/contrib/pageslts/}).
    }%
\makeatother

\bigskip

Save per page about $200\unit{ml}$~water, $2\unit{g}$~CO$_{2}$
and $2\unit{g}$~wood:\newline
Therefore please print only if this is really necessary.\newline
I do NOT think, that it is necessary to print THIS file, really\newline
(at least not after this page)!

\bigskip

\noindent The page (\verb|\thepage|): \thepage \newline

\noindent Last page's name (LastPage): \pageref{LastPage}

\newpage

\noindent The page (\verb|\thepage|): \thepage \newline

\noindent Last page's name (LastPage): \pageref{LastPage}

\newpage

\noindent The page (\verb|\thepage|): \thepage \newline

\noindent Last page's name (LastPage): \pageref{LastPage}

\newpage

\section*{The End}

\noindent The page (\verb|\thepage|): \thepage \newline

\noindent Last page's name (LastPage): \pageref{LastPage}
\end{document}
\endinput
%%
%% End of file `lastpage-example.tex'.
